Concurrent Programming with Harmony

### 2020-08-10 book/paper.tex
Latest

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\begin{verbatim}
while True:
    pass;
;
\end{verbatim}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{@cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert the following code:
\begin{code}
\begin{tabbing}
XX\=XXX\=XXX\kill
\textbf{while} \texttt{True}: \\
\> \textbf{pass}; \\
;
\end{tabbing}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{problems}

@@

In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The process continues to be polite, waiting in a \textbf{while} loop
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 be at the start of the \textbf{while} loop.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

>>>>

However, mutual exclusion by itself is easy to ensure.
For example, we could insert the following code to enter the
critical section:
\begin{code}
\texttt{await False;}
\end{code}
This code will surely prevent two or more processes from being
at label \texttt{@cs} at the same time.
But it does so by preventing \emph{any} process from reaching
the critical section.
We clearly need another property besides mutual exclusion.

@@

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert
\begin{code}
\texttt{await False;}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{problems}

@@

In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The process continues to be polite, waiting
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 be at the start of the \textbf{await} statement.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will stop waiting and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the negation of the condition on the \textbf{await}
statement for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.
